A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
seccion variables

Two set of variables has been selected to explain the diary incidence ratio in each Spanish province. The first, are climatic variables (temperature, humidity and sunshine) and its impact has been argued in the introduction section. The second, is a set of specific characteristics of each province. The objective is include in our model some control variables (socioeconomic characteristics) to capture the probable heterogeneity of each spatial unit. All those variables are constant along the temporal period analized in this research. In this subsetion we present some evidences of the expected relationship between the four control variables considered an the incidence ratio.

The structure of population is a key variable on the incidence. Several papers related the relationship between a fast trasnmision of COVID-19 and population density (e.g. cumulative incidence in urban areas like NYC) and therefore a positive sign will be expected. Note that our analysis is at provincial level - not a urban level - and therefore is necessary to rethinking the relationship. In Spain the population is concentrate (with som exception) in the mediterranean coast (@goerlich2013population), in general with high temperature and high humidity. So, positive or negative sign will be expected in the model.

The percentaje of older adults (over 65) is the second variable related with the population structure include like control variable. Unlike the population density the older people in Spain is concentrate in the core of the Iberian Peninsula (see figure \ref{plotfer1}).

The per capita GDP also can generate differences between provinces. No direct relationship can be observed between the GDP and the ratio of incidence but the GDP is a proxy of inobservable information. This varaible has been included in similar papers (@wang2020high). For low DGPpc is more difficult maintain social distance (sing +); High DGPpc -> more economic activity -> es mas difícl mantener el confinamiento.

@@

```{r plotfer1, echo=FALSE,  warning=FALSE, fig.cap="\\label{Fig-quantile} Spatial distribution of Spanish density population", cache=TRUE}
q <- quantile(provinces_spain$Density)
provinces_spain$Quantile<- as.factor((provinces_spain$Density > q[2]) + (provinces_spain$Density > q[3]) +(provinces_spain$Density >= q[4]) + 1)
p1 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
q <- quantile(provinces_spain$Older)
provinces_spain$Quantile<- as.factor((provinces_spain$Older > q[2]) + (provinces_spain$Older > q[3]) +(provinces_spain$Older >= q[4]) + 1)
p2 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
gridExtra::grid.arrange(p1,p2,nrow=1)
```

>>>>

Two set of variables has been selected to explain the daily incidence ratio in each Spanish province. The first, are climatic variables (temperature, humidity and sunshine) and its impact has been argued in the introduction section. The second, is a set of specific characteristics of each province. The objective is include in our model some control variables (socioeconomic characteristics) to capture the probable heterogeneity of each spatial unit. All those variables are time constant along the temporal period analized in this research but have high spatial variability. In this subsetion we present some evidences of the expected relationship between the four control variables and the incidence ratio.

The structure of population is a key variable on the incidence. Several papers related the relationship between a fast trasnmision of COVID-19 and population density (e.g. cumulative incidence in urban areas like NYC) and therefore a positive sign will be expected. Note that our analysis is at provincial level - not a urban level - and therefore is necessary to rethinking the relationship. In Spain the population is concentrate (with some exception) in the mediterranean coast (@goerlich2013population), in general with high temperature and high humidity. So, positive or negative sign will be expected in the model.

The percentaje of older adults (over 65) is the second variable related with the population structure include like control variable. Unlike the population density the older people in Spain is concentrate in the core of the Iberian Peninsula (see figure \ref{plotfer1}). Our hypothesis is that in the provinces with more older adults the social contact is lower (@prem2017projecting) and therefore the a negative correlation will be expected. This hypotesis is based in the idea that the agee-structured social contact matrix in Spain is similar to Italy (see @singh2020age Fig 1 pag 3)

The per capita GDP (GDPpc) also can generate differences between provinces. No direct relationship can be observed between the GDP and the ratio of incidence but our hypothesis is that the GDPpc is a proxy of inobservable information. By example, provinces with high DGPpc have more transit of foreing people This variable has been included in similar papers (@wang2020high). For low DGPpc is more difficult maintain social distance (sing +); High DGPpc -> more economic activity -> es mas difícl mantener el confinamiento.

@@

```{r plotfer1, echo=FALSE,  warning=FALSE, fig.cap="\\label{Fig-quantile} Spatial distribution of Spanish density population", cache=TRUE}
q <- quantile(provinces_spain$Density)
provinces_spain$Quantile<- as.factor((provinces_spain$Density > q[2]) + (provinces_spain$Density > q[3]) +(provinces_spain$Density >= q[4]) + 1)
p1 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
q <- quantile(provinces_spain$Older)
provinces_spain$Quantile<- as.factor((provinces_spain$Older > q[2]) + (provinces_spain$Older > q[3]) +(provinces_spain$Older >= q[4]) + 1)
p2 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
q <- quantile(provinces_spain$GDPpc)
provinces_spain$Quantile<- as.factor((provinces_spain$GDPpc > q[2]) + (provinces_spain$GDPpc > q[3]) +(provinces_spain$GDPpc >= q[4]) + 1)
p3 <- ggplot(data = provinces_spain) +
  geom_sf(aes(fill = Quantile),color = "black",size=.2) +
  theme_bw(base_size=6) +
  scale_fill_manual(values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D"))
gridExtra::grid.arrange(p1,p2,p3,nrow=2)
```
